The brightness and spatial distributions of terrestrial radio sources

### 2013-02-26 articles/Article-RFI-distribution/article-rfi-distribution.tex
Author list full and suggestions by Zarka

\author[A.~R.~Offringa et al.]{
  A.~R.~Offringa$^{1,2,3}$\thanks{E-mail:
\url{offringa@mso.anu.edu.au}},
A.~G.~de~Bruyn$^{3,4}$
and S.~Zaroubi$^3$%,
%L.~V.~E.~Koopmans$^3$,
%\newauthor
%S.~J.~Wijnholds$^4$,
%F.~B.~Abdalla$^5$,
%W.~Brouw$^{3,4}$,
%B.~Ciardi$^6$,
%I.~T.~Iliev$^7$,
%G.~J.~A.~Harker$^8$,
%\newauthor
%G.~Mellema$^8$
\\
$^1$Australian National University, RSAA Mt. Stromlo Observatory, Cotter Road, Weston Creek ACT 2611, Australia\\
$^2$ARC Centre of Excellence for All-sky Astrophysics (CAASTRO)\\
$^3$University of Groningen, Kapteyn Astronomical Institute, PO Box 800, 9700 AV Groningen, The Netherlands\\
$^4$ASTRON, PO Box 2, 7990 AA Dwingeloo, The Netherlands\\
%$^5$ UCL Department of Physics and Astronomy, London WC1E 6BT, United Kingdom\\
%$^6$Max-Planck Institute for Astrophysics, Karl-Schwarzschild-Strasse 1, 85748 Garching bei M\"unchen, Germany\\
%$^7$University of Sussex, Falmer, Brighton BN1 9QH, UK\\
%$^8$Center for Astrophysics and Space Astronomy, University of Colorado Boulder, CO 80309, USA\\
%$^9$Stockholm University, AlbaNova University Center, Stockholm Observatory, SE-106 91 Stockholm, Sweden 
}

@@

\maketitle
\label{firstpage}

@@

Next, we assume that all RFI sources have equal constant strength $I$ and follow a uniform spatial distribution in a two-dimensional plane. This is obviously a simplification, as the RFI sources are actually distributed on the surface of a sphere with different heights. Beyond some distance, the Earth will partly block the path between transmitter and receiver. Therefore, the assumption of uniformly distributed sources on a two-dimensional plane is only approximately valid. Using these assumptions, we can express the expected cumulative frequency density of sources $F_{\textrm{distance} \le r}(r)$ at distance $r$ as
\begin{equation} \label{eq:cumulative-distance-function}
F_{\textrm{distance} \le r}(r)=\rho \pi r^2,
\end{equation}
for some constant $\rho$ that represents the number of sources per unit area. In other words, we will observe $F$ sources that are at most a distance $r$ away.

@@

Converting from a subtracted term in decibels to a flux density factor $L_S$, and collecting the terms of Eq.~\eqref{eq:hata-model} that are not depending on $r$ in a single variable $\zeta$, results in
\begin{equation} \label{eq:propagation-loss-factor}
 L_S = \zeta r^{4.49 - 0.655 \log_{10} h_b},
\end{equation}
with
\begin{equation} \label{eq:zeta-definition}
 \zeta = \frac{f_c^{2.616}}{h_b^{1.382}} - 10^{6.955-\frac{1}{10} a(h_m)}.
\end{equation}
Note that according to Hata's model, the exponent of the power law in Eq.~\eqref{eq:propagation-loss-factor} depends only on the height of the transmitting antenna, i.e., it is independent of frequency, receiver height and urban density. Now, if in Eqs.~\eqref{eq:cumulative-rfi-distribution} and \eqref{eq:two-dimensional-distribution} one replaces the definition of $S(r)$ from Eq.~\eqref{eq:amplitude-fall-off-in-free-space} with one that includes the propagation effects,
\begin{equation}
 S(r) = \frac{Ig}{\zeta r^\eta},
\end{equation}
with $\eta = 4.49 - 0.655 \log_{10} h_b$, one finds the frequency density function $f_p$ that considers propagation effects,
% S(r):=I*g/(r^alpha);
% assume(g>0,I>0,S>0);
% diff(rho*%pi*r^2,S),solve(S(r)=S,r);
\begin{align} \label{eq:frequency-density-with-propagation}
f_p(S)
= & \frac{d}{dS} \left[ N - \rho \pi \left( \frac{Ig }{\zeta S} \right)^{2/\eta} \right]
= \frac{\rho 2\pi}{\eta S} \left( \frac{Ig}{\zeta S} \right)^{2/\eta}.
\end{align}
Consequently, due to non-free-space propagation effects, the observed log-log histogram is predicted to have a $-(\frac{2}{\eta}+1)$ slope. By substituting $\eta$, one finds
\begin{equation}
\textrm{slope}(h_b) = \frac{1}{0.3275 \log_{10} h_b - 2.245} - 1.
\end{equation}
\begin{figure}
\begin{center}\hspace{-1cm}\includegraphics[width=6cm]{img/transmitter-height/plot-transmitter-height-trimmed}
\caption{Effect of transmitter height on the slope of a log-log histogram. According to Hata's model, this is valid for the range 30--200 m. The trend of the slope will not continue indefinitely when increasing the height further. Instead it will converge to a $-2$ slope, which corresponds with free-space propagation.}
\label{fig:plot-transmitter-height}
\end{center}
\end{figure}
This is valid for transmitters that have a height of 30--200 m, the range over which Hata's model was defined. This yields estimated distribution slopes of $-1.57$ and $-1.67$ for $30$ m and $200$ m high transmitters respectively. In Figure~\ref{fig:plot-transmitter-height}, the slope value is plotted as a function of the transmitter height, including extrapolated values for transmitter heights down to 1 m.

\subsection{Noise contribution} \label{sec:histogram-noise}
The full measured distribution will consist of the power-law distribution combined with that of the noise and the celestial signal. For now, we will ignore the contribution of the celestial signal, as its contribution to the amplitude distribution will be minimal when observing fields without strong sources. For example, the strongest apparent source in the NCP EoR field is around 5 Jy (Yattawata et al. 2013, \textit{in press}). % TODO fill in details.
The standard deviation of the noise, however, is around 100~Jy on highest LOFAR resolutions, and will have a larger contribution on the histogram.

@@

\label{lastpage}
\section*{Acknowledgments}
LOFAR, the Low-Frequency Array designed and constructed by ASTRON, has facilities in several countries, that are owned by various parties (each with their own funding sources), and that are collectively operated by the International LOFAR Telescope (ILT) foundation under a joint scientific policy.

@@

>>>>

\author[A.~R.~Offringa et al.]{A.~R.~Offringa$^{1,2,3}$\thanks{E-mail:
\url{offringa@mso.anu.edu.au}},
A.~G.~de~Bruyn$^{4,3}$, S.~Zaroubi$^{3}$, L.~V.~E.~Koopmans$^{3}$, \newauthor
S.~J.~Wijnholds$^{4}$, F.~B.~Abdalla$^{5}$, W.~N.~Brouw$^{3,4}$, B.~Ciardi$^{6}$, I.~T.~Iliev$^{7}$, \newauthor
G.~J.~A.~Harker$^{8}$, G.~Mellema$^{9}$, G.~Bernardi$^{10}$, P.~Zarka$^{11}$, A.~Ghosh$^{3}$,\newauthor A.~Alexov$^{12}$, J.~Anderson$^{13}$, A.~Asgekar$^{4}$, I.~M.~Avruch$^{14,3}$, R.~Beck$^{13}$,\newauthor
M.~E.~Bell$^{2}$, M.~R.~Bell$^{6}$, M.~J.~Bentum$^{4}$, P.~Best$^{15}$, L.~B\^{i}rzan$^{16}$,
F.~Breitling$^{17}$, \newauthor
J.~Broderick$^{18}$, M.~Br\"uggen$^{19}$, H.~R.~Butcher$^{4,1}$, F.~de~Gasperin$^{19}$, E.~de~Geus$^{4}$,\newauthor
M.~de~Vos$^{4}$, S.~Duscha$^{4}$, J.~Eisl\"offel$^{20}$, R.~A.~Fallows$^{4}$, C.~Ferrari$^{21}$,\newauthor
W.~Frieswijk$^{4}$, M.~A.~Garrett$^{4,16}$, J.~Grie\ss{}meier$^{22}$, T.~E.~Hassall$^{18}$, A.~Horneffer$^{13}$,\newauthor
M.~Iacobelli$^{16}$, E.~Juette$^{23}$, A. ~Karastergiou$^{24}$, W.~Klijn$^{4}$, V.~I.~Kondratiev$^{4,25}$,\newauthor
M.~Kuniyoshi$^{13}$, G.~Kuper$^{4}$, J.~van~Leeuwen$^{4,26}$, M.~Loose$^{4}$, P.~Maat$^{4}$,\newauthor
G.~Macario$^{21}$, G.~Mann$^{17}$, J.~P.~McKean$^{4}$, H.~Meulman$^{4}$, M.~J.~Norden$^{4}$, 
\newauthor
E.~Orru$^{4}$, H.~Paas$^{27}$, M.~Pandey-Pommier$^{28}$, R.~Pizzo$^{4}$, A.~G.~Polatidis$^{4}$,\newauthor
D.~Rafferty$^{16}$, W.~Reich$^{13}$, R.~van~Nieuwpoort$^{4}$, H.~R\"ottgering$^{16}$, \newauthor
A.~M.~M.~Scaife$^{18}$, J.~Sluman$^{4}$, O.~Smirnov$^{29,30}$, C.~Sobey$^{13}$, M.~Tagger$^{22}$,\newauthor
Y.~Tang$^{4}$, C.~Tasse$^{11}$, S.~ter~Veen$^{31}$, C.~Toribio$^{4}$, R.~Vermeulen$^{4}$, C.~Vocks$^{17}$,\newauthor
R.~J.~van~Weeren$^{10}$, M.~W.~Wise$^{4,26}$, O.~Wucknitz$^{32,13}$
\\
$^{1}$RSAA, Australian National University, Mt Stromlo Observatory, via Cotter Road, Weston, ACT 2611, Australia \\
$^{2}$ARC Centre of Excellence for All-sky Astrophysics (CAASTRO) \\
$^{3}$Kapteyn Astronomical Institute, PO Box 800, 9700 AV Groningen, The Netherlands \\
$^{4}$Netherlands Institute for Radio Astronomy (ASTRON), Postbus 2, 7990 AA Dwingeloo, The Netherlands \\
$^{5}$UCL Department of Physics and Astronomy, London WC1E 6BT, United Kingdom \\
$^{6}$Max Planck Institute for Astrophysics, Karl Schwarzschild Str. 1, 85741 Garching, Germany \\
$^{7}$University of Sussex, Falmer, Brighton BN1 9QH, UK \\
$^{8}$Center for Astrophysics and Space Astronomy, University of Colorado Boulder, CO 80309, USA \\
$^{9}$Stockholm University, AlbaNova University Center, Stockholm Observatory, SE-106 91 Stockholm, Sweden \\
$^{10}$Harvard-Smithsonian Center for Astrophysics, 60 Garden Street, Cambridge, MA 02138, USA \\
$^{11}$LESIA, UMR CNRS 8109, Observatoire de Paris, 92195   Meudon, France \\
$^{12}$Space Telescope Science Institute, 3700 San Martin Drive, Baltimore, MD 21218, USA \\
$^{13}$Max-Planck-Institut f\"{u}r Radioastronomie, Auf dem H\"ugel 69, 53121 Bonn, Germany \\
$^{14}$SRON Netherlands Insitute for Space Research, Sorbonnelaan 2, 3584 CA, Utrecht, The Netherlands \\
$^{15}$Institute for Astronomy, University of Edinburgh, Royal Observatory of Edinburgh, Blackford Hill, Edinburgh EH9 3HJ, UK \\
$^{16}$Leiden Observatory, Leiden University, PO Box 9513, 2300 RA Leiden, The Netherlands \\
$^{17}$Leibniz-Institut f\"{u}r Astrophysik Potsdam (AIP), An der Sternwarte 16, 14482 Potsdam, Germany \\
$^{18}$School of Physics and Astronomy, University of Southampton, Southampton, SO17 1BJ, UK \\
$^{19}$University of Hamburg, Gojenbergsweg 112, 21029 Hamburg, Germany \\
$^{20}$Th\"{u}ringer Landessternwarte, Sternwarte 5, D-07778 Tautenburg, Germany \\
$^{21}$Laboratoire Lagrange, UMR7293, Universit\`{e} de Nice Sophia-Antipolis, CNRS, Observatoire de la C\'{o}te d'Azur, 06300 Nice, France \\
$^{22}$Laboratoire de Physique et Chimie de l' Environnement et de l' Espace, LPC2E UMR 7328 CNRS, 45071 Orl\'{e}ans Cedex 02, France \\
$^{23}$Astronomisches Institut der Ruhr-Universit\"{a}t Bochum, Universitaetsstrasse 150, 44780 Bochum, Germany \\
$^{24}$Astrophysics, University of Oxford, Denys Wilkinson Building, Keble Road, Oxford OX1 3RH \\
$^{25}$Astro Space Center of the Lebedev Physical Institute, Profsoyuznaya str. 84/32, Moscow 117997, Russia \\
$^{26}$Astronomical Institute 'Anton Pannekoek', University of Amsterdam, Postbus 94249, 1090 GE Amsterdam, The Netherlands \\
$^{27}$Center for Information Technology (CIT), University of Groningen, The Netherlands \\
$^{28}$Centre de Recherche Astrophysique de Lyon, Observatoire de Lyon, 9 av Charles Andr\'{e}, 69561 Saint Genis Laval Cedex, France \\
$^{29}$Centre for Radio Astronomy Techniques \& Technologies (RATT), Department of Physics and Elelctronics, Rhodes University, PO Box 94, Grahamstown 6140, South Africa \\
$^{30}$SKA South Africa, 3rd Floor, The Park, Park Road, Pinelands, 7405, South Africa \\
$^{31}$Department of Astrophysics/IMAPP, Radboud University Nijmegen, P.O. Box 9010, 6500 GL Nijmegen, The Netherlands \\
$^{32}$Argelander-Institut f\"{u}r Astronomie, University of Bonn, Auf dem H\"{u}gel 71, 53121, Bonn, Germany
}

@@

\label{firstpage}
\maketitle

@@

Next, we assume that all RFI sources have equal constant strength $I$ and follow a uniform spatial distribution in the local two-dimensional horizontal plane. This is obviously a simplification, as the RFI sources are actually distributed on the surface of a sphere with different heights. Beyond some distance, the Earth will partly block the path between transmitter and receiver. Therefore, the assumption of uniformly distributed sources on a two-dimensional plane is only approximately valid. Using these assumptions, we can express the expected cumulative frequency density of sources $F_{\textrm{distance} \le r}(r)$ at distance $r$ as
\begin{equation} \label{eq:cumulative-distance-function}
F_{\textrm{distance} \le r}(r)=\rho \pi r^2,
\end{equation}
for some constant $\rho$ that represents the number of sources per unit area. In other words, we will observe $F$ sources that are at most a distance $r$ away.

@@

Converting from a subtracted term in decibels to a flux density factor $L_S$, and collecting the terms of Eq.~\eqref{eq:hata-model} that are not depending on $r$ in a single variable $\zeta$, results in
\begin{equation} \label{eq:propagation-loss-factor}
 L_S = \zeta r^{4.49 - 0.655 \log_{10} h_b},
\end{equation}
with
\begin{equation} \label{eq:zeta-definition}
 \zeta = \frac{f_c^{2.616}}{h_b^{1.382}} - 10^{6.955-\frac{1}{10} a(h_m)}.
\end{equation}
Note that according to Hata's model, the exponent of the power law in Eq.~\eqref{eq:propagation-loss-factor} depends only on the height of the transmitting antenna, i.e., it is independent of frequency, receiver height and urban density. Now, if in Eqs.~\eqref{eq:cumulative-rfi-distribution} and \eqref{eq:two-dimensional-distribution} one replaces the definition of $S(r)$ from Eq.~\eqref{eq:amplitude-fall-off-in-free-space} with one that includes the propagation effects,
\begin{equation}
 S(r) = \frac{Ig}{\zeta r^\eta},
\end{equation}
with $\eta = 4.49 - 0.655 \log_{10} h_b$, one finds the frequency density function $f_p$ that considers propagation effects,
% S(r):=I*g/(r^alpha);
% assume(g>0,I>0,S>0);
% diff(rho*%pi*r^2,S),solve(S(r)=S,r);
\begin{align} \label{eq:frequency-density-with-propagation}
f_p(S)
= & \frac{d}{dS} \left[ N - \rho \pi \left( \frac{Ig }{\zeta S} \right)^{2/\eta} \right]
= \frac{\rho 2\pi}{\eta S} \left( \frac{Ig}{\zeta S} \right)^{2/\eta}.
\end{align}
Consequently, due to non-free-space propagation effects, the observed log-log histogram is predicted to have a $-(\frac{2}{\eta}+1)$ slope. By substituting $\eta$, one finds
\begin{equation}
\textrm{slope}(h_b) = \frac{1}{0.3275 \log_{10} h_b - 2.245} - 1.
\end{equation}
\begin{figure}
\begin{center}\hspace{-1cm}\includegraphics[width=6cm]{img/transmitter-height/plot-transmitter-height-trimmed}
\caption{Effect of transmitter height on the slope of a log-log histogram. According to Hata's model, this is valid for the range 30--200 m. The trend of the slope will not continue indefinitely when increasing the height further. Instead it will converge to a $-2$ slope, which corresponds to free-space propagation.}
\label{fig:plot-transmitter-height}
\end{center}
\end{figure}
This is valid for transmitters that have a height of 30--200 m, the range over which Hata's model was defined. This yields estimated distribution slopes of $-1.57$ and $-1.67$ for $30$ m and $200$ m high transmitters respectively. In Figure~\ref{fig:plot-transmitter-height}, the slope value is plotted as a function of the transmitter height, including extrapolated values for transmitter heights down to 1 m.

\subsection{Thermal noise contribution} \label{sec:histogram-noise}
The full measured distribution will consist of the power-law distribution combined with that of the thermal noise and the celestial signal. For now, we will ignore the contribution of the celestial signal, as its contribution to the amplitude distribution will be minimal when observing fields without strong sources. For example, the strongest apparent source in the NCP EoR field is around 5 Jy (Yattawata et al. 2013, \textit{in press}). % TODO fill in details.
The standard deviation of the noise, however, is around 100~Jy on highest LOFAR resolutions, and will have a larger contribution on the histogram.

@@

\section*{Acknowledgments}
LOFAR, the Low-Frequency Array designed and constructed by ASTRON, has facilities in several countries, that are owned by various parties (each with their own funding sources), and that are collectively operated by the International LOFAR Telescope (ILT) foundation under a joint scientific policy.

@@

\label{lastpage}
